Validation of marker gene expression

### 2025-09-12 analyses/cell-type-consensus/template-notebooks/marker-gene-validation.Rmd
update validation notebook template

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
  library(patchwork)
  library(data.table)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)
```

@@

# path to results directory
results_dir <- file.path(module_base, "..", "..", "data", "current", "results", "cell-type-consensus", params$project_id)
```

@@

```{r, warning=FALSE}
# get consensus results 
consensus_results_files <- list.files(results_dir, pattern = "_processed_consensus-cell-types\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)

# get gene expression files 
gene_exp_files <- list.files(results_dir, pattern = "_processed_marker-gene-expression\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)

@@

```{r, message=FALSE, warning=FALSE}
# read in consensus and gene expression results 
# use vroom and data.table to try and be more memory-efficient
consensus_dt <- consensus_results_files |> 
  purrr::map(fread) |> 
  data.table::rbindlist(fill = TRUE, use.names = TRUE)

gene_exp_dt <- gene_exp_files |> 
  purrr::map(fread) |> 
  data.table::rbindlist(fill = TRUE, use.names = TRUE)

# read in validation markers as data.tables
markers_dt <- fread(validation_markers_file)

validation_groups_dt <- fread(validation_groups_file) |> 
  # rename final assigned group to avoid conflicts when merging in marker gene expression 
  # we want to separate the marker gene group from the actual cell type annotation
  dplyr::select(consensus_annotation, broad_celltype_group = validation_group_annotation)
```

```{r, results='asis'}
# check for if the project is a cell line project or not 
# explicitly skip project 24 since the current processed data doesn't have the updated combination of PDX and cell line
is_cell_line <- all(consensus_dt$sample_type == "cell line") | params$project_id == "SCPCP000024"

@@

```{r, eval=!is_cell_line, message=FALSE}
# combine into one data frame that has all marker genes
# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
consensus_dt <- consensus_dt |> 
  # add in broad cell type group which is used for plotting
  # groups similar cell types together 
  dplyr::left_join(validation_groups_dt, by = "consensus_annotation") |> 
  dplyr::left_join(gene_exp_dt, by = c("barcodes", "library_id")) |> 
  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
  # account for the same gene being present in multiple cell types 
  dplyr::left_join(markers_dt, by = "ensembl_gene_id", relationship = "many-to-many") |> 
  dplyr::mutate(detected = logcounts > 0)

# remove the extra data table 
rm(gene_exp_dt)
```

@@

```{r, eval=!is_cell_line}
# prep for plots 
# get total number of cells per final annotation group 
total_cells_df <- consensus_dt |> 
  dplyr::select(library_id, barcodes, broad_celltype_group) |> 
  dplyr::distinct() |> 
  dplyr::count(broad_celltype_group, name = "total_cells") 
```

```{r, eval=!is_cell_line}
# table with one row per unique broad cell type/ marker gene combination 
# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
# do this for every possible marker gene/ validation group assignment 
# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
group_stats_df <- consensus_dt |> 
  # for each assigned cell type/marker gene combo get total detected and mean expression
  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
  dplyr::group_by(broad_celltype_group, ensembl_gene_id, validation_group_annotation) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |> 
  # add in validation group for marker genes
  # this includes all possible marker genes and all possible validation group assignments 
  dplyr::left_join(markers_dt, by = c("ensembl_gene_id", "validation_group_annotation"), relationship = "many-to-many") |>
  # now get the mean expression/ mean percentage across all marker genes for a given validation group
  # here the broad_celltype_group is the final assigned annotation for that group of cells 
  # the validation_group_annotation refers to the cell type that marker gene is associated with 
  dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
  dplyr::mutate(
    # calculate mean expression/detected across all markers for a specific group 
    all_markers_mean_exp = mean(mean_exp),
    all_markers_detected_count = mean(detected_count)
  ) |>  # add total cells
  dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |> 
  # for plotting we're only going to look at any cell types with > 50 cells otherwise these plots can get wild 
  dplyr::filter(total_cells > 50) |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # get total percent expressed
    percent_exp = (detected_count/total_cells) * 100,
    all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100, 
    # account for NA/unknowns and set axes order
    broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |> 
      factor(levels =  c(unique(markers_dt$validation_group_annotation), "unknown"))
  )

# no longer need this and it takes up space 
rm(consensus_dt)

@@

# assign colors to use in annotation bars 
celltype_colors <- palette.colors(palette = "alphabet")[1:length(celltype_groups)]
names(celltype_colors) <- celltype_groups
```

@@

```{r, eval=!is_cell_line}
# filter markers to those that are actually relevant 
# we will only plot the marker genes for cell types that are part of the assigned broad validation group for this project
# we don't care about plotting marker genes for cell types that aren't present here 
# note that we will use this for both the dotplot and the heatmap
filtered_markers_df <- markers_dt |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% group_stats_df$gene_symbol)

@@

```{r, eval=!is_cell_line, fig.height=10, fig.width=15}
# filter out low expressed genes
dotplot_df <- unique_gene_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 5) |> 
  dplyr::arrange(broad_celltype_group) |> 
  # add a label for the plot 
  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})"))) |> 
  # remove marker genes that aren't present in final annotations and set x axis order 
  dplyr::filter(gene_symbol %in% marker_gene_order) |> 
  dplyr::mutate(
    # set orders of gene symbol and validation groups 
    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
    validation_group_annotation = factor(validation_group_annotation, levels = celltype_order)
  )

@@

# add annotation bar aligning marker genes with validation group 
color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = validation_group_annotation)) + 
  geom_tile() + 
  scale_fill_manual(values = celltype_colors, breaks = levels(dotplot_df$validation_group_annotation)) +
  ggmap::theme_nothing() +
  theme(legend.position = "bottom") +
  labs(fill = "")

dotplot + color_bar +
  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1)) 
```

>>>>

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
  library(patchwork)
})

# Set default ggplot theme
theme_set(
  theme_classic()
)

# Use DuckDB for dplyr functions
suppressMessages(duckplyr::methods_overwrite())
```

@@

# path to results directory
results_dir <- file.path(module_base, "results", "cell-type-consensus", params$project_id)
```

@@

```{r, warning=FALSE}
# get consensus results 
consensus_results_files <- list.files(results_dir, pattern = "_consensus-cell-types\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)

# get gene expression files 
gene_exp_files <- list.files(results_dir, pattern = "_marker-gene-expression\\.tsv\\.gz$", recursive = TRUE, full.names = TRUE)

@@

```{r, message=FALSE, warning=FALSE}
# read in validation markers as data.tables
markers_df <- duckplyr::read_csv_duckdb(validation_markers_file) |> 
  duckplyr::as_duckdb_tibble()

validation_groups_df <- duckplyr::read_csv_duckdb(validation_groups_file) |> 
  # rename final assigned group to avoid conflicts when merging in marker gene expression 
  # we want to separate the marker gene group from the actual cell type annotation
  dplyr::select(consensus_annotation, broad_celltype_group = validation_group_annotation) |> 
  duckplyr::as_duckdb_tibble()

# read in files directly to duckdb tables
  # specify all varchar for consensus to avoid parsing error
consensus_df <- duckplyr::read_csv_duckdb(consensus_results_files, options = list(sep = "\t", union_by_name = TRUE))

gene_exp_df <- duckplyr::read_csv_duckdb(gene_exp_files, options = list(sep = "\t", union_by_name = TRUE))
```

```{r, results='asis'}
# check for if the project is a cell line project or not 
# explicitly skip project 24 since the current processed data doesn't have the updated combination of PDX and cell line
is_cell_line <- all(consensus_df$sample_type == "cell line") | params$project_id == "SCPCP000024"

@@

```{r, eval=!is_cell_line, message=FALSE}
# combine into one data frame that has all marker genes
# one row for each cell/marker gene combination (total rows= total cells x total marker genes)
# includes original consensus annotation, braod cell type group, marker gene id, logcounts, validation group annotation, information regarding expression of marker gene in CellMarker tissues 
consensus_df <- consensus_df |> 
  # add in broad cell type group which is used for plotting
  # groups similar cell types together 
  dplyr::left_join(validation_groups_df, by = "consensus_annotation") |> 
  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |>
  # add marker gene information (associated validation group annotation, gene observed count, percent tissues)
  # account for the same gene being present in multiple cell types 
  # we only want genes that are in both and throw out extra consensus genes 
  dplyr::left_join(markers_df, by = "ensembl_gene_id", relationship = "many-to-many") |> 
  dplyr::mutate(detected = logcounts > 0)
```

@@

```{r, eval=!is_cell_line}
# prep for plots 
# get total number of cells per final annotation group 
total_cells_df <- consensus_df |> 
  dplyr::select(library_id, barcodes, broad_celltype_group) |> 
  dplyr::distinct() |> 
  dplyr::count(broad_celltype_group, name = "total_cells") 
```

```{r, eval=!is_cell_line}
# table with one row per unique broad cell type/ marker gene combination 
# first all cells in with the same broad_celltype_group (determined based on consensus_annotation) are grouped together
# then get the mean gene expression and total percentage of cells that express each marker gene across all cells in that group
# do this for every possible marker gene/ validation group assignment 
# second we calculate the mean expression and mean percentage of all marker genes in a given validation group (this value is used only in the second section of the report)
group_stats_df <- consensus_df |> 
  # for each assigned cell type/marker gene combo get total detected and mean expression
  # group by both broad group and validation group to account for genes that are expressed in more than one cell type
  dplyr::summarize(
    .by = c("broad_celltype_group", "ensembl_gene_id", "validation_group_annotation"),
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |> 
  # add in validation group for marker genes
  # this includes all possible marker genes and all possible validation group assignments 
  dplyr::left_join(markers_df, by = c("ensembl_gene_id", "validation_group_annotation"), relationship = "many-to-many") |>
  # now get the mean expression/ mean percentage across all marker genes for a given validation group
  # here the broad_celltype_group is the final assigned annotation for that group of cells 
  # the validation_group_annotation refers to the cell type that marker gene is associated with 
  dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
  dplyr::mutate(
    # calculate mean expression/detected across all markers for a specific group 
    all_markers_mean_exp = mean(mean_exp),
    all_markers_detected_count = mean(detected_count)
  ) |>  # add total cells
  dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |> 
  # for plotting we're only going to look at any cell types with > 50 cells otherwise these plots can get wild 
  dplyr::filter(total_cells > 50) |> 
  dplyr::mutate(
      # get total percent expressed
      percent_exp = (detected_count / total_cells) * 100,
      all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100,
      # account for NA/unknowns and set axes order
      broad_celltype_group = tidyr::replace_na(broad_celltype_group, "unknown") |>
        factor(levels = c(unique(markers_df$validation_group_annotation), "unknown"))
    )

# no longer need this and it takes up space 
rm(consensus_df)

@@

# assign colors to use in annotation bars 
celltype_colors <-  colorRampPalette(RColorBrewer::brewer.pal(12, "Paired"))(length(celltype_groups))
names(celltype_colors) <- celltype_groups
```

@@

```{r, eval=!is_cell_line}
# filter markers to those that are actually relevant 
# we will only plot the marker genes for cell types that are part of the assigned broad validation group for this project
# we don't care about plotting marker genes for cell types that aren't present here 
# note that we will use this for both the dotplot and the heatmap
filtered_markers_df <- markers_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% group_stats_df$gene_symbol)

@@

```{r, eval=!is_cell_line, fig.height=15, fig.width=25}
# filter out low expressed genes
dotplot_df <- unique_gene_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 5) |> 
  dplyr::arrange(broad_celltype_group) |> 
  # add a label for the plot 
  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})"))) |> 
  # remove marker genes that aren't present in final annotations and set x axis order 
  dplyr::filter(gene_symbol %in% marker_gene_order) |> 
  dplyr::mutate(
    # set orders of gene symbol and validation groups 
    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
    validation_group_annotation = factor(validation_group_annotation, levels = celltype_order)
  )

@@

# add annotation bar aligning marker genes with validation group 
color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = validation_group_annotation)) +
  geom_tile() +
  facet_grid(cols = vars(validation_group_annotation), scales = "free", space = "free") +
  scale_fill_manual(values = celltype_colors, breaks = levels(dotplot_df$validation_group_annotation)) +
  ggmap::theme_nothing() +
  theme(
    strip.background = element_rect(fill = "transparent", color = NA),
    strip.text.x = element_text(angle = 90, hjust = 0, vjust = 0.5, size = 12),
    strip.placement = "outside",
    legend.position = "none",
    panel.spacing = unit(0.5, "lines"),
    strip.clip = "off"
  ) +
  labs(fill = "")

color_bar / dotplot +
  patchwork::plot_layout(ncol = 1, heights = c(0.1, 4))
```
